\documentclass[12pt,a4paper,openright]{report}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\usepackage{geometry}
\geometry{hmargin=2.5cm,vmargin=2.5cm}
\usepackage{xcolor}
\usepackage{tikz}

\usepackage{listings}
\lstset{
    language=Python,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\bfseries,
    stringstyle=\color{red},
    commentstyle=\color{green!60!black}\itshape,
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    frame=single,
    breaklines=true
}

\usepackage[colorlinks=true,
            linkcolor=blue!60!black,
            urlcolor=blue!60!black,
            citecolor=blue!60!black,
            bookmarks=true,
            bookmarksnumbered=true,
            pdftitle={Fundamentals of NLP},
            pdfauthor={Aymen Ben Brik}]{hyperref}

% Theorem environments (continuous numbering)
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\newtheorem{property}{Property}[section]

\title{
    \vspace*{2cm}
    {\Huge \textbf{Fundamentals of}} \\[0.3cm]
    {\Huge \textbf{Natural Language Processing}} \\[1cm]
    {\Large Course material} \\[0.3cm]
    {\large M1 GAMMA \quad \&\quad L2 LMAD --- Data Science}
}
\author{Aymen Ben Brik}
\date{\today}

\begin{document}

% --- Title page ---
\begin{titlepage}
\maketitle
\vfill
\begin{center}
\begin{tabular}{ll}
\textbf{Course} & Fundamentals of Natural Language Processing \\
\textbf{Audience} & M1 GAMMA \\
                  & L2 LMAD --- Data Science \\
\textbf{Author} & Aymen Ben Brik \\
\textbf{Repository} & \url{https://github.com/Aymenbenbrik/FundamentalsNLP} \\
\end{tabular}
\end{center}
\vfill
\end{titlepage}

% --- Foreword ---
\chapter*{Foreword}
\addcontentsline{toc}{chapter}{Foreword}

This document is the unified support for an introductory Natural Language Processing course aimed at first-year M1 (GAMMA) and second-year L2 (LMAD Data Science) students. It assumes basic Python programming, working knowledge of machine learning and linear algebra (vectors, dot products, distances), and a willingness to read mathematical definitions.

\paragraph{Pedagogy.} Each chapter follows a unified structure: a concrete \emph{motivation}, an \emph{approach by problem}, formal \emph{definitions and theorems} with proofs where appropriate, \emph{worked examples} of ascending difficulty, integrated \emph{Python code} (\texttt{scikit-learn}, \texttt{NLTK}, \texttt{gensim}, \texttt{spaCy}, \texttt{transformers}), and \emph{exercises}. Many examples and exercises are framed in a Tunisian or multilingual context to ensure relevance to the target audience.

\paragraph{Companion material.} Four Jupyter lab notebooks accompany the chapters in the \texttt{Notebooks/} folder of the GitHub repository:
\begin{itemize}
    \item \texttt{lab01-tokenization.ipynb} --- whitespace, regex, BPE from scratch, BERT vs GPT-2 tokenizers.
    \item \texttt{lab02-bow-tfidf-classifier.ipynb} --- Bag-of-Words, TF--IDF, logistic-regression classifier with feature interpretation.
    \item \texttt{lab03-word2vec-embeddings.ipynb} --- skip-gram / CBOW, similarity, analogies, PCA visualization.
    \item \texttt{lab04-evaluation-metrics.ipynb} --- classification metrics, ROC / PR curves, perplexity, BLEU, ROUGE.
\end{itemize}

\paragraph{Learning objectives.} By the end of this course, the student should be able to:
\begin{enumerate}
    \item Define NLP and identify its three core challenges (ambiguity, context, variability).
    \item Recognize the major NLP tasks and their real-world applications.
    \item Situate NLP within the broader AI / ML / DL landscape and identify the dominant learning paradigms.
    \item Distinguish discriminative from generative modelling approaches.
    \item Trace the historical evolution of NLP from rule-based systems to large language models.
    \item Apply tokenization strategies appropriate to a given task and language.
    \item Build sparse (BoW, TF--IDF) and dense (Word2Vec, GloVe, FastText) text representations.
    \item Compute and interpret cosine, dot-product and Euclidean similarities.
    \item Evaluate NLP systems with task-appropriate metrics (Precision, Recall, F$_1$, perplexity, BLEU, ROUGE) and recognise their pitfalls.
\end{enumerate}

% --- Table of contents ---
\tableofcontents

% --- Chapters ---
\input{ch1}
\input{ch2}

\end{document}
